Stage-Splitting a Modal Language

### 2014-09-24 paper/semantics.tex
added figure and outlined a new example/intro (filling out at home)

\section{The \lang\ Language}
\label{sec:semantics}

\input {figures/grammars}
\input {figures/statics}

In this section, we describe \lang, a simple two-stage language
for which we define and analyze stage splitting.

% \lang\ is largely based off the
% language \lamCircle\ by \cite{davies96}.  
% which was used as a model for partial evaluation.  
% Our version, despite a slightly different semantics, can be used for the same purpose.
% The details of the partial evaluation view are shown in \cref{sec:partialeval}.

\lang\ is essentially two copies of a typed lambda calculus with
products, sums, and isorecursive types.%
\footnote{It is possible for different stages to have different sets of
features, but for simplicity we do not consider this.}
Every valid expression has both a type and a stage, either \bbone~or \bbtwo,
expressing in which copy of the language the term resides. Intuitively, the
stage of a term expresses \emph{when} to evaluate it---all stage-\bbone\
subexpressions are evaluated before stage-\bbtwo\ ones.

>>>>

\section {The \lang\ Language}

\begin{figure*}
\begin{subfigure}{0.5\textwidth}
\begin{lstlisting} 
datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) 
  : (int*list*list) =
  case unroll l of 
    Empty => (0,Empty, Empty) 
  | Cons (h,t) =>
      let (s,left,right) = partition (p,t) in
      if h<p 
      then (s+1,Cons(h,left),right)
      else (s,left,Cons(h,right))

fun qs (l : list, i : int) : int = 
  case l of
    Empty => next {0}
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
        case compare i n of
          LT => qs left i
        | EQ => h
        | GT => qs right (i-n-1)
\end{lstlisting}
\caption{Unstaged Code}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\begin{lstlisting} 
datatype list = Empty | Cons of int * list

fun partition (p : int, l : list) = ...
fun qs (l : list, i: $int) : $int = 
  case l of
    Empty => next {0}
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
      next{
        let n = hold{n} in
          case compare prev{i} n of
            LT => prev {qs left i}
          | EQ => hold {h}
          | GT => prev {qs right next{prev{i}-n-1}}
      }	
\end{lstlisting}
\caption{Staged Code}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\begin{lstlisting} 
datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) = ...

datatype tree = Branch of int * int * tree * tree
                | Leaf

fun qs1 (l : list) : tree =
  case l of
    Empty => Leaf
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
      Branch (n, h, qs1 left i, qs1 right i)

fun qs2 (p : tree, i : int) : int = 
  case unroll p of
    Leaf => 0
  | Branch (n,h,p1,p2) => 
      case compare i n of
        LT => qs2 (p1,i)
      | EQ => h
      | GT => qs2 (p2,i-n-1)
\end{lstlisting}
\caption{Split Code}
\end{subfigure}
\caption{Caption place holder}
\end{figure*}

Consider the quickselect algorithm in [the figure],
which finds the $i$th largest element of a list.
It operates by inspecting the head of the list, 
partitioning the rest of the list by the head,
comparing the size of the partition to the desired index,
and the recursively selecting on the correct sublist.
If the index was outside valid bounds, 0 is returned.
[runtime]

It has been observed that [dependency...].

We can then leverage this dependency to split quickselect into two other functions [one precomputes all splits... build tree... other consumes].
[We've shown this in figure...]
[we like it as two functions because... multiple calls... efficient]

The goal of our work is to do this transformation automatically.
[first we need a language to ... unambiguously and clearly specify staging]
[ideally that language looks similar to the original term]
[The type system for this language should be enough to prove that splitting will work out.]

[to this end, we present \lang]
The main idea of \lang\ is that terms defined in it are 
split across two {\em stages}, called \bbone\ and \bbtwo.
If part of the term is in \bbone, then it will end up in the precomputation, else it will end up in the residual.
[information can flow from the stage \bbone\ portions of the term to the stage \bbtwo\ portions,
but never from \bbtwo\ to \bbone.  this property is required if we want to be able to split the term]

[Check out the example, now staged all fancy.]

This staged version contains three new constructs: 
$\next$ and $\prev$ at the term level, and $\fut$ at the type level.
We can read $\next$ and $\prev$ as indications to 
execute the code in the following block at stage \bbtwo\ and \bbone, respectively.
The type constructor $\fut$ is a way for us to talk at stage \bbone\ about things that will not be available until stage \bbtwo.
Thus, ${\tt fexp} : \fut\rmint \times \rmint \to \fut\rmint$ 
can be read as a function that takes a base at stage \bbtwo\ and a power at stage \bbone, 
and returns the exponent at stage \bbtwo.

[$\fut$ acts in a sanitary way]

\section{Statics and Dynamics}
\label{sec:semantics}

\input {figures/grammars}
\input {figures/statics}
%
\footnote{It is possible for different stages to have different sets of
features, but for simplicity we do not consider this.}
Every valid expression has both a type and a stage, either \bbone~or \bbtwo,
expressing in which copy of the language the term resides. Intuitively, the
stage of a term expresses \emph{when} to evaluate it---all stage-\bbone\
subexpressions are evaluated before stage-\bbtwo\ ones.
